## Packages for this section
```{r, eval=F}
library(lubridate)
```

## Dates
- Dates represented on computers as “days since an origin”,
typically Jan 1, 1970, with a negative date being before the origin:
```{r}
dates=c("1970-01-01","2007-09-04","1931-08-05")
(d=as.Date(dates))
```

```{r}
as.numeric(d)
```

## Doing arithmetic with dates

- Dates are "actually" numbers, so can add and subtract: xxx

```{r}
d+30
d[2]-d[3]
```

## Reading in dates from a file
- `read_csv` and the others can guess that you have dates, if you
format them as year-month-day, like column 1 of this `.csv`:

```
date,status,dunno
2011-08-03,hello,August 3 2011
2011-11-15,still here,November 15 2011
2012-02-01,goodbye,February 1 2012
```

- Then read them in:

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/c32/mydates.csv"
ddd=read_csv(my_url)
```

- read_csv guessed that the 1st column is dates, but not 3rd.

## The data as read in

```{r}
ddd
```

## Dates in R
- Preceding shows that dates should be stored as text in format
yyyy-mm-dd (ISO standard).
- To deal with dates in other formats, use package `lubridate` xxx and
convert. For example, dates in US format with month first:
```{r}
usdates=c("05/27/2012","01/03/2016","12/31/2015")
mdy(usdates)
```

- For UK-format dates with month second, one of these dates is legit,
but the other two make no sense:

```{r}
dmy(usdates)
```

## That last column in our data frame

- That is month, day, year, so:

```{r}
(d4=ddd %>% mutate(date2=mdy(dunno)))
```

## Are they really the same? xxx

- Column date2 was correctly converted from column dunno: xxx

```{r}
d4 %>% mutate(equal=identical(date,date2))
```

- The two columns of dates are all the same.

## Making dates from pieces
Starting from this file:

```
year month day
1970 1 1
2007 9 4
1940 4 15
```


```{r}
my_url="http://www.utsc.utoronto.ca/~butler/c32/pieces.txt"
dates0=read_delim(my_url," ")
```

## Making some dates xxx

```{r}
dates0 %>%
unite(dates,day,month,year) %>%
mutate(d=dmy(dates)) -> newdates
```


## The results

```{r}
newdates
```

- `unite` glues things together with an underscore between them (if you
don’t specify anything else). Syntax: first thing is new column to be
created, other columns are what to make it out of.
- `unite` makes the original variable columns year, month, day
*disappear*.
- The column `dates` is text, while `d` is a real date.

## Extracting information from dates xxx

```{r}
newdates %>% 
  mutate(
    mon=month(d),
    day=day(d),
    weekday=wday(d, label=T)
  )
```

## Dates and times
- Standard format for times is to put the time after the date, hours,
minutes, seconds:
```{r}
(dd=tibble(text=c("1970-01-01 07:50:01","2007-09-04 15:30:00",
"1940-04-15 06:45:10","2016-02-10 12:26:40")))
```

## Converting text to date-times: xxx

- Then get from this text using `ymd_hms`:
```{r}
dd %>% mutate(dt=ymd_hms(text))
```

- Default timezone is “Universal Coordinated Time”. Change it via `tz=`
and the name of a timezone:

```{r}
(dd %>% mutate(dt=ymd_hms(dd,tz="America/Toronto")) -> dd)
```

## Extracting time parts
- As you would expect: xxx
```{r}
dd %>% mutate(h=hour(dt),
              sec=second(dt),
              min=minute(dt),
              zone=tz(dt))
```

## Same times, but different time zone: xxx

```{r}
dd %>% mutate(oz=with_tz(dt,"Australia/Sydney"))
```

## How long between date-times?
- We may need to calculate the time between two events. For example,
these are the dates and times that some patients were admitted to
and discharged from a hospital:

```
admit,discharge
1981-12-10 22:00:00,1982-01-03 14:00:00
2014-03-07 14:00:00,2014-03-08 09:30:00
2016-08-31 21:00:00,2016-09-02 17:00:00
```

## Do they get read in as date-times? xxx

- These ought to get read in and converted to date-times:

```{r}
my_url="http://www.utsc.utoronto.ca/~butler/c32/hospital.csv"
stays=read_csv(my_url)
```

-  and so it proves.

## Subtracting the date-times

- In the obvious way, this gets us an answer:
```{r}
stays %>% mutate(stay=discharge-admit)
```

- Number of hours; hard to interpret. 

## Days

- Fractional number of days would
be better:

```{r}
stays %>% mutate(stay=(discharge-admit)/ddays(1))
```

## Comments
- Date-times are stored internally as seconds-since-something, so that
subtracting two of them will give, internally, a number of seconds.
- Just subtracting the date-times is displayed as a time (in units that R
chooses for us).
- Functions `ddays(1)`, `dminutes(1)` etc. will give number of seconds
in a day or a minute, thus dividing by them will give (fractional) days,
minutes etc.
- This idea useful for calculating time from a start point until an event
happens (in this case, a patient being discharged from hospital).

