## Packages for this section
```{r, eval=F}
library(tidyverse)
library(lubridate)
```

## Exact time intervals

We previously got fractional days (of stays in hospital):

```{r}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/hospital.csv"
stays <- read_csv(my_url)
stays %>% mutate(stay_days = (discharge - admit) / ddays(1))
```

but what if we wanted days, hours and minutes?

## Intervals

\footnotesize

```{r}
stays %>% mutate(stay = admit %--% discharge)
```

\normalsize

- These are called *intervals*: they have a start point and an end point.

## Periods

To work out the exact length of an interval, in human units, turn it into a `period`:

```{r}
stays %>% mutate(stay = as.period(admit %--% discharge))
```

A period is exact as long as it has a start and an end (accounting for daylight savings, leap years etc).

## Completed days

Take `day` of the periods:

```{r}
stays %>% mutate(stay = as.period(admit %--% discharge)) %>% 
  mutate(days_of_stay = day(stay))
```

## Completed hours

- Not quite what you think:

```{r}
stays %>% mutate(stay = as.period(admit %--% discharge)) %>% 
  mutate(hours_of_stay = hour(stay))
```

- These are completed hours *within* days. 
- To get total hours, count each day as 24 hours also:

```{r}
stays %>% mutate(stay = as.period(admit %--% discharge)) %>% 
  mutate(hours_of_stay = hour(stay) + 24*day(stay))
```

## Sometimes it matters

- Days and hours are always the same length (as a number of seconds).
- Months and years are not always the same length:
  - months have different numbers of days
  - years can be leap years or not
  - the actual length of 2 months depends which 2 months you are looking at:
  
```{r}
tribble(
  ~start, ~end,
  ymd("2020-01-15"), ymd("2020-03-15"),
  ymd("2020-07-15"), ymd("2020-09-15")
) %>% mutate(per = as.period(start %--% end)) %>% 
  mutate(secs = (end - start)/ dseconds(1))
```

- so dividing by eg `dyears` could be problematic.

## Manchester United

Sometime in December 2019 or January 2020, I downloaded some information about the players that were then in the squad of the famous Manchester United Football (soccer) Club. We are going to use the players' ages (as given) to figure out exactly when the download happened.

```{r}
my_url <- "http://ritsokiguess.site/STAD29/manu.csv"
read_csv(my_url) %>% select(name, date_of_birth, age) -> man_united
man_united
```

## Fractional years

- guess a date for when I downloaded this,
- turn date of birth into a date, 
- work out fractional years:

```{r}
dl_date <- ymd("2020-01-05")
man_united %>% 
  mutate(date_of_birth = dmy(date_of_birth)) %>% 
  mutate(years = (dl_date - date_of_birth) / dyears(1))
```

## See which ages were gotten wrong

`floor` gets the integer part of a decimal number:

```{r}
man_united %>% 
  mutate(date_of_birth = dmy(date_of_birth)) %>% 
  mutate(years = (dl_date - date_of_birth) / dyears(1)) %>% 
  filter(floor(years) != age)
```

- These three players had a birthday between the date I downloaded the data and the date I guessed.

```{r}
man_united %>% 
  mutate(date_of_birth = dmy(date_of_birth)) %>% 
  mutate(calc_age = as.period(date_of_birth %--% dl_date)) %>% 
  mutate(yrs = year(calc_age)) %>% 
  filter(age != yrs)
```

## going back

- what if I go back to December 1?

```{r}
dl_date <- ymd("2019-12-01")
man_united %>% 
  mutate(date_of_birth = dmy(date_of_birth)) %>% 
  mutate(calc_age = as.period(date_of_birth %--% dl_date)) %>% 
  mutate(yrs = year(calc_age)) %>% 
  filter(age != yrs)
```

## notes

- fractional years get all the ages right
- so there is no advantage to using a period