---
title: "STAC32: Applications of Statistical Methods"
author: "Lecture notes"
header-includes:
   - \usepackage{multicol}
output: 
  beamer_presentation:
    latex_engine: lualatex
    slide_level: 2
    df_print: kable
    theme: "AnnArbor"
    colortheme: "dove"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height = 5)
# knitr::opts_chunk$set(echo = FALSE)
options(width=55)
```

# Course outline

# Running R

```{r child="running_R.Rmd"}
```

# Reading in data

```{r child="readfile_R.Rmd"}
```

# Making graphs

```{r, child="graphs_R.Rmd"}
```

# Numerical summaries: more detailed

```{r, child="numsum_R.Rmd"}
```

# Statistical Inference

```{r, child="inference_R.Rmd"}
```

# Writing reports

```{r, child="reports_R.Rmd"}
```


# Tidying and organizing data

```{r, child="tidying_R.Rmd"}
```

xxx from here later 

# the end of the notes

# template

# The beginning

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```


# the end