---
title: "STAC32: Applications of Statistical Methods"
author: "Lecture notes"
header-includes:
   - \usepackage{multicol}
output: 
  beamer_presentation:
    latex_engine: lualatex
    slide_level: 2
    df_print: kable
    theme: "AnnArbor"
    colortheme: "dove"
---

```{r child="preliminaries.Rmd", echo=F}
```

xxx put in a "packages for this section" each time

# Course outline xxx

# Running R

```{r child="running_R.Rmd"}
```

# Reading in data

```{r child="readfile_R.Rmd"}
```

# Making graphs

```{r, child="graphs_R.Rmd"}
```

# Numerical summaries: more detailed

```{r, child="numsum_R.Rmd"}
```

# Statistical Inference

```{r, child="inference_R.Rmd"}
```

# Writing reports

```{r, child="reports_R.Rmd"}
```


# Tidying and organizing data

```{r, child="tidying_R.Rmd"}
```

# Case study 1: the windmill data

```{r, child="windmill.Rmd"}
```


# Case study 3: Asphalt

```{r, child="asphalt.Rmd"}
```



# Regression with categorical variables

```{r, child="with_categ_R.Rmd"}
```

# Functions

```{r, child="functions.Rmd"}
```

# Dates and times 

```{r, child="dates_and_times.Rmd"}
```

# Miscellanea

```{r, child="miscellanea.Rmd"}
```


# Vector and matrix algebra (for C33)

```{r, child="vector_matrix.Rmd"}
```

# piece on bootstrap

# piece on rstan

# the end of the notes

# template

# The beginning

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for autAhoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.

## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Output

```{r cars, echo = TRUE}
summary(cars)
```

## Slide with Plot

```{r pressure}
plot(pressure)
```


# the end